---
title: "Activity Rhythm Analysis in Drosophila"
output:
  html_document:
    df_print: paged
  pdf_document: default
---

1. Importing required packages
```{r}
#Packages to import, store and manipulate behavioral data
library(behavr, damr)
#For Graphical Visualization of behavioral data. Extension of ggplot2. Additionally Pre-processes data.
library(ggetho) 
#For behavioral data analysis
library(sleepr, zeitgebr)


```


2. Loading dams_sample data available as a part of damr
```{r}
getwd()
data(dams_sample) 
#Make a copy to manipulate data
dt <- copy(dams_sample) 
summary(dt)
print(head(dt[meta=TRUE]))
```
32 individual fies under DD.
Period_group metavariable can be either a “short”,”long“ or wild-type (“wt”) period


3. Marking regime transitions. Example: LD12 -> DD
    t = 0 now means “ZT0 of the transition day”
```{r}
#DT[, baseline_days:= 3, meta=TRUE] #Days
#dt[, t := t - days(xmv(baseline_days))] #To seconds
```


4. Data Enrichment
    DAM data only has variables t and activity. Activity is the number of beam crosses over a time bin (e.g. one minute)
```{r}
head(dt[, moving := activity > 0])
# moving is TRUE when and only when activity > 0, and FALSE otherwise
```


5. Exploratory Tile Plot
    To spot outliers
    Subjective day: Greys, Subjective Night: Black
```{r}
ggetho(dt, aes(z=activity)) +
      stat_ld_annotations(ld_colours = c("grey", "black"))+
      stat_tile_etho() 
```

    
6. Dealing with dead animals - 5 & 10 
    Removing data after death based on long bouts of inactivity
```{r}
dt_curated <- curate_dead_animals(dt)
summary(dt_curated)
```


7. Tile plot with outliers removed
```{r}
ggetho(dt_curated, aes(z=activity)) +
      stat_ld_annotations(ld_colours = c("grey", "black"))+
      stat_tile_etho() #variable of interest as a colour with pixel intensity
```


8. Changing naming conventions to simpler ones
    Assigning a number for each individual fly
```{r}
head(dt_curated[, uid := 1 : .N, meta=T])
head(dt_curated[, .(id, uid) ,meta=T]) #map uid to id
```


9. Double Plotted Actograms using multiplot = 2
```{r}
ggetho(dt_curated, aes(z = moving), multiplot = 2) + 
    stat_bar_tile_etho() + #bar height to show the amount of movement
    facet_wrap( ~ period_group + uid, ncol = 8, labeller = label_wrap_gen(multi_line=FALSE)) 
        #split graph by given uid and period_group
        #Graph names in a single line
```


10. Periodograms
  a. Computing the periodicity of the free running clock of multiple individuals
      Representation of the density (i.e. power) of a signal at different periods (or frequencies). 
      In addition, a periodogram associates to each pair of power-period a significance level
```{r}
per_xsq_dt <- periodogram(activity, 
                        dt_curated,
                        FUN = chi_sq_periodogram) #ac_periodogram – An autocorrelation based method
                                                  #ls_periodogram – Lomb-Scargle algorithm
head(per_xsq_dt)
```
Additional Information on original functions: lomb::lsp; xsp::chiSqPeriodogram; stats::acf 


  b.Peaks finding
      Finds a maximum of three peaks, which are sorted by their power (relative to the significance thershold)
      Insignificant peaks are left out
```{r}
per_xsq_dt <- find_peaks(per_xsq_dt)
head(per_xsq_dt)
```
      Annotates our data by adding a column named "peak". 
      Whenever the row corresponds to a peak, it puts a number and NA otherwise. 
  
      
  c. Visualisation
      i. PLot of all periodograms
```{r}
ggperio(per_xsq_dt) + geom_line(aes(group = id, colour=period_group))
```
  
  
    ii. Periodograms of the individual flies
```{r}
ggperio(per_xsq_dt) + 
  geom_line(aes(group = id, colour = period_group)) +
  geom_peak(col = "black") +
  geom_line(aes(y = signif_threshold)) +
  facet_wrap(~ uid, ncol = 8)
ggperio(per_xsq_dt) + 
  geom_line(aes(group = id, colour = period_group)) + #Group by uid
  geom_peak(peak_rank = 1:2, col = "black") +         #show peak values for the I & II most significant peak values
  geom_line(aes(y = signif_threshold)) +              #Draw line of significance
  facet_wrap(~ uid, ncol = 8)
```
  
  
    iii. Population average periodogram
```{r}
ggperio(per_xsq_dt, aes(y = power - signif_threshold,
                  colour=period_group)) + 
                  stat_pop_etho()
```
    
    
  d. Extracting Peak Values of 1
      Significant differences in peak periodicity vs genotype
```{r}
summary_dt <- rejoin(per_xsq_dt[peak==1]) #Merge metadata and summary statistics 
head(summary_dt)
```
  
  
  e. Distribution of periods
      Size of the points is proportional to the relative power of the peak discovered
```{r}
ggplot(summary_dt, aes(period_group, period, fill= period_group)) + 
        geom_boxplot(outlier.colour = NA) +
        geom_jitter(aes(size=power -  signif_threshold), alpha=.5) +
        scale_y_hours(name = "Period") 
```
     
      
      f. Ascertaining statistically significant differences in period between each pair of groups
```{r}
pairwise.wilcox.test(summary_dt$period, summary_dt$period_group )
```
      

      